% AI in Genomics — Project Writeup
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{enumitem}
\geometry{margin=1in}

\title{Modeling Protein Sequence Adaptation to Environmental Factors\\
       Using Gold Database Samples and UniProt Sequences}
\author{NYU AI in Genomics Project}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This document describes the data acquisition, enrichment, feature engineering,
modeling and evaluation plan for predicting protein sequence variation as a
function of environmental factors. Samples (with locality information) from the
Gold database will be linked to UniProt protein sequences. Locality coordinates
will be enriched with environmental covariates (climate, soil, oceanographic,
and anthropogenic metrics). The enriched dataset will be used to train
conditional sequence prediction models that learn how environmental changes
affect protein sequences. This writeup is designed to be kept up-to-date as the
project progresses and to serve as the authoritative project PDF.
\end{abstract}

\section{Background and Motivation}
Many extremophiles and environment-adapted organisms exhibit sequence features
correlated with environmental pressures (temperature, salinity, pH, radiation,
etc.). By systematically linking curated sample metadata (location, substrate,
temperature recorded at sampling) from the Gold dataset to protein sequences in
UniProt, and enriching each sample with high-resolution environmental covariates
derived from the sampling coordinates, we can build models that predict
sequence-level changes in response to environmental factor perturbations.

\section{Objectives}
\begin{enumerate}[nosep]
  \item Gather Gold database samples with precise geocoordinates and linked
    specimen identifiers.
  \item Retrieve corresponding protein sequences (or sequence identifiers)
    from UniProt for each sample where available.
  \item Enrich each sample with environmental covariates derived from the
    geographic coordinates.
  \item Construct feature pipelines to combine environment data and sample
    metadata, and construct labels from sequence representations.
  \item Train conditional sequence prediction models to predict sequence
    variation under controlled changes to environmental inputs.
  \item Evaluate model fidelity with sequence-based metrics and ecological
    plausibility checks.
\end{enumerate}

\section{Data Sources and Retrieval}
\subsection{Gold database (samples and metadata)}
\begin{itemize}[nosep]
    \item Primary source for sample metadata: sample identifier, sampling date, latitude/longitude, sampled substrate, measurement metadata (if available).
    \item Retrieval: use Gold API or downloadable TSV/CSV export. Filter for records with valid coordinates and sample-level identifiers that can be linked to sequence records.
\end{itemize}

\subsection{UniProt (sequences)}
\begin{itemize}[nosep]
    \item UniProt entries are retrieved using UniProt accession numbers/TAX ID or by mapping sample identifiers (where the Gold entry contains links to sequence accessions or organism identifiers).
    \item Retrieval methods: UniProt REST API, batch queries, or local UniProt downloads. Use accession mapping endpoints or TAX ID to translate between gene or sample IDs and protein sequences when needed.
\end{itemize}

\subsection{Linking strategy}
\begin{itemize}[nosep]
    \item Exact match: if Gold provides UniProt accession or EMBL/GenBank accessions, use them directly to fetch canonical protein sequences.
    \item Taxonomic + gene matching: when direct accessions are missing, use taxon and locus/gene information to identify candidate UniProt entries; record confidence and provenance for each mapping.
\end{itemize}

\section{Environmental Enrichment (Work in Progress)}
Given a latitude/longitude for each sample, enrich with the following covariates. Prefer open datasets with programmatic access and clear spatial resolution metadata.

\subsection{Static and long-term climatologies}
\begin{itemize}[nosep]
    \item WorldClim / CHELSA: long-term monthly climatologies (temperature, precipitation, bioclimatic variables).
    \item SoilGrids: soil pH, organic carbon, texture classes.
\end{itemize}

\subsection{Short-term / time-specific weather and radiation}
\begin{itemize}[nosep]
    \item NASA POWER / ERA5 (or ERA5-Land): surface temperature, radiation, humidity for the sampling date (if available) or for a representative window.
\end{itemize}

\subsection{Oceanographic variables (for marine samples)}
\begin{itemize}[nosep]
    \item Marine datasets: sea-surface temperature (SST), salinity, pH proxies, chlorophyll-a (from MODIS/Aqua or OISST), and depth/bathymetry.
\end{itemize}

\subsection{Anthropogenic and land-use covariates}
\begin{itemize}[nosep]
    \item Human footprint indices, land cover classes, proximity to built-up areas and agricultural extent (from MODIS/ESA Land Cover products).
\end{itemize}

\subsection{Implementation notes for enrichment}
\begin{itemize}[nosep]
    \item For raster datasets (WorldClim, SoilGrids, etc.) use a spatial sampling library (rasterio / xarray / rioxarray in Python) to extract point values at sample coordinates. Handle coordinate reference systems explicitly.
    \item For model-ready time-series (e.g., NASA POWER), align the sampling date to the environmental record and compute summary statistics (mean, min, max, anomalies) over relevant windows (sampling day, 7-day, 30-day, seasonal).
    \item Record data provenance and resolution (e.g., WorldClim v2 2.5 arc-min).
\end{itemize}

\section{Feature Engineering (Work in Progress)}
\subsection{Environmental feature vector}
\begin{itemize}[nosep]
    \item Numeric variables: temperature metrics, precipitation, salinity, pH, radiation, soil organic matter, elevation, distance-to-coast, etc.
    \item Categorical variables: land cover class, substrate type (rock, sediment, biofilm), biome.
    \item Normalization: scale continuous variables (z-score or quantile scaling) and encode categorical variables with one-hot or learned embeddings.
\end{itemize}

\subsection{Sequence label representation}
We consider multiple label formulations depending on modeling goals:
\begin{enumerate}[nosep]
  \item Residue-level prediction: predict the amino-acid at each aligned
    position. Requires multiple-sequence alignment (MSA) and fixed-length
    positional labels (useful for single-protein families).
  \item K-mer / signature prediction: predict distribution over k-mer counts
    or sequence-derived summary statistics (GC-equivalent, hydrophobicity
    indexes).
  \item Embedding/regression targets: map sequences to continuous embeddings
    (from pre-trained protein language models) and predict embedding vectors as
    a regression conditioned on environmental features.
  \item Generative sequence model: learn conditional sequence generator
    (transformer, conditional VAE) that outputs full amino-acid sequences given
    environmental features.
\end{enumerate}

\subsection{Choice trade-offs}
\begin{itemize}[nosep]
    \item Alignment-based residue prediction is interpretable but requires clean MSAs and homologous sequences.
    \item Embedding/regression approaches are flexible for diverse sequence lengths and can leverage pre-trained protein language models (e.g., ESM, ProtBert).
\end{itemize}

\section{Modeling Approaches (Work in Progress)}
\subsection{Conditional discriminative models}
\begin{itemize}[nosep]
    \item Predictors for residue or embedding targets: feed-forward networks, gradient-boosted trees (for aggregated labels), or sequence-to-label CNN/LSTM models when inputs include local sequence context.
\end{itemize}

\subsection{Conditional generative models}
Sequence generation conditioned on environment using
\begin{itemize}[nosep]
    \item Conditional Transformer decoder (conditioning via concatenated tokens or cross-attention to environment embeddings).
    \item Conditional VAE: latent space captures sequence variation and decoder reconstructs sequences given environment vectors.
\end{itemize}

\subsection{Model inputs and conditioning}
\begin{itemize}[nosep]
    \item Simple conditioning: concatenate normalized environment vector to each token embedding or to the decoder's initial states.
    \item Cross-attention: allow decoder layers to attend to a learned environment embedding sequence.
\end{itemize}

\section{Training Pipeline and Reproducibility}
\subsection{Data pipeline}
Steps:
\begin{itemize}[nosep]
    \item[1] Ingest Gold sample rows with coordinates; \item[2] Map to UniProt sequences;
    \item[3] Enrich coordinates with environmental rasters/APIs;
    \item[4] Precompute sequence labels and embeddings;
    \item[5] Construct training/validation splits stratified by geography, taxon, and sampling time where appropriate.
\end{itemize}

\subsection{Implementation notes}
\begin{itemize}[nosep]
    \item Organize code under `src/' with modular components: `data/' (downloaders, mappers), `features/' (environment extraction, normalization), `models/', and `experiments/' (training harness, hyperparameter configs).
    \item Save provenance metadata (source, raster name, extraction timestamp) in a companion JSON or Parquet table for each sample.
\end{itemize}

\subsection{Reproducibility}
\begin{itemize}[nosep]
    \item Use pinned package versions in `requirements.txt' or `environment.yml'.
    \item Provide a `README.md' with commands to reproduce the dataset enrichment and training. Store random seeds and checkpoint model weights in `checkpoints/'.
\end{itemize}

\section{Evaluation and Validation}
\subsection{Quantitative metrics}
\begin{itemize}[nosep]
    \item Per-residue accuracy, per-sequence edit distance (Levenshtein), BLEU-style scores adapted for amino-acids, and embedding-space cosine similarity.
    \item Calibration metrics for probabilistic models (Brier score, negative log-likelihood).
\end{itemize}

\subsection{Ecological and biological validation}
\begin{itemize}[nosep]
    \item Compare predicted sequences' biophysical properties (e.g., predicted thermostability, isoelectric point, hydrophobicity) against expected trends for environmental changes.
    \item Case studies: hold out geographically or environmentally distinct regions and test whether predicted sequence adjustments are biologically plausible.
\end{itemize}

\section{Experiments and Expected Results}
\begin{itemize}[nosep]
  \item Baseline: predict sequence embeddings with a linear model from
    environmental covariates; measure embedding reconstruction error.
  \item Generative baseline: conditional LSTM/Transformer that generates
    short protein regions given environment vectors.
  \item Ablations: remove subsets of environmental factors (temperature,
    salinity) to quantify their marginal effects on sequence predictions.
  \item Sensitivity analysis: perturb input covariates and measure sequence
    changes to estimate model responsiveness.
\end{itemize}

\section{Data Privacy, Ethics, and Limitations}
\begin{itemize}[nosep]
    \item Track and respect any access restrictions or licensing for Gold or UniProt subsets. If geolocation data are sensitive, consider obfuscation or aggregation to larger spatial units.
    \item Limitations: observational confounding (correlated environmental variables, unobserved sampling biases), phylogenetic dependence among samples, and the challenge of causally attributing sequence changes to environment.
\end{itemize}

% \section{Practical How-To: Running the Pipeline}
% \begin{enumerate}[nosep]
%   \item Prepare environment: create a Python virtualenv and install
%     dependencies from `requirements.txt`.
%   \item Fetch Gold sample exports and save to `data/raw/gold_samples.csv`.
%   \item Run `src/uniprot_and_gold_data.py` to map Gold entries to UniProt and
%     download sequences to `data/processed/sequences/`.
%   \item Run `src/environmental_enrichment.py` to extract raster/APIs values
%     to `data/processed/env_features.parquet`.
%   \item Run training: `python src/train.py --config experiments/config.yaml`.
% \end{enumerate}

% \section{Keeping the Writeup Updated}
% - This file `writeup.tex` should be edited as experiments progress. To rebuild
%   the PDF locally:
% \begin{verbatim}
% pdflatex writeup.tex
% bibtex writeup (if using citations)
% pdflatex writeup.tex
% pdflatex writeup.tex
% \end{verbatim}

\section{Next Steps}
\begin{itemize}[nosep]
  \item Implement automated mapping between Gold and UniProt and store a
    provenance table.
  \item Prototype environmental extraction for a small sample set (10--100
    points) to validate raster/API choices.
  \item Build a baseline embedding regression and evaluate.
\end{itemize}

\section{Appendix: Recommended packages and tools}
\begin{itemize}[nosep]
  \item Python: `requests`, `pandas`, `geopandas`, `rasterio`, `xarray`,
    `rioxarray`, `scikit-learn`, `pytorch`/`tensorflow`, `biopython`.
  \item Data sources: UniProt REST API, WorldClim, CHELSA, SoilGrids,
    NASA POWER, ERA5, MODIS/Ocean products.
\end{itemize}

\vfill
\noindent Last updated: \today.
\end{document}
